---
title: "CHiCAGO Vignette"
author: "Jonathan Cairns, Paula Freire-Pritchett, Mikhail Spivakov"
date: 17th November, 2015
VignetteIndexEntry: 
output:
  BiocStyle::html_document:
    toc: true
bibliography: Chicago.bib
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{CHiCAGO Vignette}
%\VignetteDepends{}
%\VignetteKeywords{Chicago}
%\VignettePackage{Chicago}
-->
```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

## Introduction

CHiCAGO is a method for detecting statistically significant interaction events in Capture HiC data. This vignette will walk you through a typical CHiCAGO analysis. 

A typical Chicago job for two biological replicates of CHi-C data takes 2-3 h wall-clock time (including sample pre-processing from bam files) and uses 50G RAM.

NOTE: A wrapper to perform this type of analysis, called *runChicago.R*, is provided as part of *chicagoTools*, which is available from our [Bitbucket repository](https://bitbucket.org/chicagoTeam/chicago/src). Refer to the *chicagoTools* README for more information.

The statistical foundations of CHiCAGO will be presented in a separate paper - a preprint is currently available [@ChicagoPreprint]. Briefly, CHiCAGO uses a convolution noise model accounting for both 'Brownian' (distance-dependent) and 'technical' noise. It borrows information across interactions (with appropriate normalisation) to estimate these noise components separately on different subsets of data. CHiCAGO then uses a p-value weighting procedure based on the expected true positive rates at different distance ranges (estimated from data), with scores representing soft-thresholded -log weighted p-values. The score threshold of 5 is a suggested stringent score threshold for calling significant interactions. 

WARNING: The data set used in this tutorial comes from the package *PCHiCdata*. This package contains small parts (two chromosomes each) of published Promoter Capture HiC data sets in mouse ESCs [@Schoenfelder2015] and GM12878 cells, derived from human LCLs [@Mifsud2015] - note that both papers used a different interaction-calling algorithm and we are only reusing raw data from them. Do not use any of these sample input data for purposes other than training. 

In this vignette, we use the GM12878 data [@Mifsud2015]:

```{r}
library(PCHiCdata)
```

## Input files required

Before you start, you will need: 

1. Five restriction map information files ("design files"):

- Restriction map file (.rmap) - a bed file containing coordinates of the restriction fragments. By default, 4 columns: chr, start, end, fragmentID.
- Bait map file (.baitmap) - a bed file containing coordinates of the baited restriction fragments, and their associated annotations. By default, 5 columns: chr, start, end, fragmentID, geneName.
- *nperbin* file (.npb), *nbaitsperbin* file (.nbpb), *proxOE* file (.poe) - Precompute these tables from the .rmap and .baitmap files, using the Python scripts from *chicagoTools* at our [Bitbucket repository](https://bitbucket.org/chicagoTeam/chicago/src): ``countNperBin.py``, ``countNbaitsPerBin.py`` and ``getProxOE.py``. Refer to the ``chicagoTools`` README file for more details.

We recommend that you put all five of these files into the same directory (that we refer to as designDir). An example of a valid design folder, for a two-chromosome sample of the GM12878 data used in this vignette, is provided in the PCHiCdata package, as follows.

```{r}
dataPath <- system.file("extdata", package="PCHiCdata")
testDesignDir <- file.path(dataPath, "hg19TestDesign")
dir(testDesignDir)
```

2. You will also need input data files. These should be in CHiCAGO input format, *.chinput*. *.chinput* files can be obtained from aligned Capture HiC BAM files by running ``bam2chicago.sh``, available as part of ``chicagoTools``. (To obtain BAM files from raw fastq files, run them through a HiC alignment & QC pipeline such as [HiCUP](http://www.bioinformatics.babraham.ac.uk/projects/hicup/).

Example *.chinput* files are provided in the PCHiCdata package, as follows:

```{r}
testDataPath <- file.path(dataPath, "GMchinputFiles")
dir(testDataPath)

files <- c(
    file.path(testDataPath, "GM_rep1.chinput"),
    file.path(testDataPath, "GM_rep2.chinput"),
    file.path(testDataPath, "GM_rep3.chinput")
  )
```

OPTIONAL: The data set in this vignette requires some additional custom settings, to ensure that the vignette compiles in a reasonable time:

```{r}
settingsFile <- file.path(system.file("extdata", package="PCHiCdata"),
                          "sGM12878Settings", "sGM12878.settingsFile")
```

Omit this step unless you know which settings you wish to alter.

## Example workflow

We run CHiCAGO on the test data as follows. First, we create a blank ``chicagoData`` object, and we tell it where the design files are. For this example, we also provide the optional settings file - in your analysis, you can omit the ``settingsFile`` argument.

```{r, message=FALSE}
library(Chicago)

cd <- setExperiment(designDir = testDesignDir, settingsFile = settingsFile)
```

The properties of ``chicagoData`` objects are discussed more in [The chicagoData object](#cd).

Next, we read in the input data files:

```{r, message=FALSE}
cd <- readAndMerge(files=files, cd=cd)
```

Finally, we run the pipeline with ``chicagoPipeline()``:

```{r, eval=FALSE}
cd <- chicagoPipeline(cd)
```

## Output plots

``chicagoPipeline()`` produces a number of plots. You can save these to disk by setting the ``outprefix`` argument in ``chicagoPipeline()``.

The plots are as follows:

1. Brownian noise other end factors: The adjustment made to the mean Brownian noise count, based on the class of the other end. ("tlb" refers to the number of trans reads the other end has, in total. "B2B" stands for a "bait-to-bait" interaction.)
2. Technical noise estimates: The mean number of technical noise reads expected for other ends and baits, respectively, in certain classes. These classes, displayed on the x axis, again refer to the number of trans reads that the other end has.
3. Distance function: The mean number of Brownian noise reads expected for an ``average`` bait, as a function of distance, plotted on a log-log scale.

```{r, echo=FALSE, message=FALSE}
cd <- chicagoPipeline(cd)
```

## Output files

Two main output methods are provided. Here, we discuss the first: exporting to disk. However, it is also possible to export to a GenomeInteractions object, as described in [Further downstream analysis](#further-downstream-analysis).

You can export the results to disk, using ``exportResults()``. (If you use *runChicago.R*, the files appear in ./\<results-folder\>/data). By default, the function outputs three different output file formats:

```{r}
outputDirectory <- tempdir()
exportResults(cd, file.path(outputDirectory, "vignetteOutput"))
```

Each called interaction is assigned a score that represents how strong CHiCAGO believes the interaction is (formally, it is based on -log(adjusted P-value)). Thus, a larger score represents a stronger interaction. In each case, the score threshold of 5 is applied.

Summary of output files:

### ibed format (ends with ...ibed):

```{r echo=FALSE}

a <- read.table(file.path(outputDirectory, "vignetteOutput.ibed"), header=TRUE)
head(a)

```

- each row represents an interaction
- first four columns give information about the chromosome, start, end and name of the bait fragment
- next four columns give information about the chromosome, start, end and name of the other end that interacts with the bait fragment
- N\_reads is the number of reads
- score is as defined above

### seqmonk format (ends with ...seqmonk.txt):

```{r echo=FALSE}

a <- read.table(file.path(outputDirectory, "vignetteOutput_seqmonk.txt"), header=FALSE)
head(a)

```


- Can be read by [seqmonk](http://www.bioinformatics.babraham.ac.uk/projects/seqmonk/).
- An interaction is represented by two rows: the first row is the bait, the second the other end. Thus, the file alternates: bait1, otherEnd1, bait2, otherEnd2, ...
- Columns are: chromosome, start, end, name, number of reads, interaction score (see above)

### washU\_text format (ends with ...washU\_text.txt):

```{r echo=FALSE}

a <- read.table(file.path(outputDirectory, "vignetteOutput_washU_text.txt"), header=FALSE)
head(a)

```


- Can be read by the [WashU browser](http://epigenomegateway.wustl.edu)
- Upload via the "Got text files instead? Upload them from your computer" link.
- Note - Advanced users may wish to export to washU\_track format instead. See the help page for ``exportResults()``.

For bait-to-bait interactions, the interaction can be tested either way round (i.e. either fragment can be considered the "bait"). In most output formats, both of these tests are preserved. The exception is washU output, where these scores are consolidated by taking the maximum.

## Visualising interactions

The ``plotBaits()`` function can be used to plot the raw read counts versus linear distance from bait for either specific or random baits, labelling significant interactions in a different colour. By default, 16 random baits are plotted, with interactions within 1 Mb from bait passing the threshold of 5 shown in red and those passing the more lenient threshold of 3 shown in blue.

```{r eval=FALSE}

plottedBaitIDs <- plotBaits(cd, n=6)

```

```{r echo=FALSE, fig.height=10}

invisible(plotBaits(cd, baits=c(415839, 404491, 425847, 417632, 409335, 414114)))

```


## Peak enrichment for features

``peakEnrichment4Features()`` tests the hypothesis that other ends in the CHiCAGO output are enriched for genomic features of interest - for example, histone marks associated with enhancers. We find out how many overlaps are expected under the null hypothesis (i.e. that there is no enrichment) by shuffling the other ends around in the genome, while preserving the overall distribution of distances over which interactions span.

You will need additional files to perform this analysis - namely, a .bed file for each feature. We include ChIP-seq data from the ENCODE consortium [@encode_integrated_2012], also restricted to chr20 and chr21. (Data accession numbers: Bernstein lab GSM733752, GSM733772, GSM733708, GSM733664, GSM733771, GSM733758)

First, we find the folder that contains the features, and construct a list of the features to use:

```{r}
featuresFolder <- file.path(dataPath, "GMfeatures")
dir(featuresFolder)

featuresFile <- file.path(featuresFolder, "featuresGM.txt")
featuresTable <- read.delim(featuresFile, header=FALSE, as.is=TRUE)
featuresList <- as.list(featuresTable$V2)
names(featuresList) <- featuresTable$V1
featuresList
```

Next, we feed this information into the ``peakEnrichment4Features()`` function.

As part of the analysis, ``peakEnrichment4Features()`` takes a distance range (by default, the full distance range over which interactions are observed), and divides it into some number of bins. We must select the number of bins; here, we choose that number to ensure that the bin size is approximately 10kb. If the defaults are changed, a different number of bins is more appropriate. See ``?peakEnrichment4Features`` for more information.

```{r, message=FALSE, fig.width=12, fig.height=7}
no_bins <- ceiling(max(abs(intData(cd)$distSign), na.rm = TRUE)/1e4)

enrichmentResults <- peakEnrichment4Features(cd, folder=featuresFolder, list_frag=featuresList,
                                             no_bins=no_bins, sample_number=100)
```

Note the plot produced by this function. For each feature type, the yellow bar represents the number of features that overlap with interaction other ends. The blue bar represents what would be expected by chance, with a 95\% confidence interval for the mean number of overlaps plotted. If the yellow bar lies outside of this interval, we reject the null hypothesis, thus concluding that there is enrichment/depletion of that feature.

The information displayed in the plot is also returned in tabular form:

```{r}
enrichmentResults
```

## Further downstream analysis

We can perform further downstream analysis in R or Bioconductor, using functionality from the `r Biocpkg("GenomicInteractions")` package. First, we export the significant interactions into a GenomicInteractions object:

```{r, message=FALSE}
library(GenomicInteractions)
library(GenomicRanges)
gi <- exportToGI(cd)
```

From here, we can pass the CHiCAGO results through to other Bioconductor functionality. In the following example, we find out which other ends overlap with the H3K4me1 enhancer mark, using ENCODE data. We use `r Biocpkg("AnnotationHub")` to fetch a relevant enhancer mark track from the ENCODE project:

```{r, message=FALSE}
library(AnnotationHub)
ah <- AnnotationHub()
hs <- query(ah, c("GRanges", "EncodeDCC", "Homo sapiens", "H3k4me1"))
enhancerTrack <- hs[["AH23254"]]
```

Next, we use the ``anchorTwo()`` function to extract the other end locations from the GenomicInteractions object (``anchorOne()`` would give us the bait locations instead). Note that in this particular instance, the ``seqlevels()`` also need to be changed before performing the comparison, adding "chr" to make them match those of the annotation.

```{r}
otherEnds <- anchorTwo(gi)
otherEnds <- renameSeqlevels(otherEnds, c("chr20","chr21"))
```

Finally, we look at which other ends overlap the enhancer marks:

```{r}
findOverlaps(otherEnds, enhancerTrack)
```

Further note that the annotation's genome version should match that of the promoter capture data, namely hg19:
```{r}
hs["AH23254"]$genome
```

## The chicagoData object {#cd}

In the above workflows, *cd* is a *chicagoData* object. It contains three elements:

* ``intData(cd)`` is a `r CRANpkg("data.table")` (note: not a *data.frame*) that contains information about fragment pairs.
* ``settings(cd)`` is a list of settings, usually set with the setExperiment() function.
* ``params(cd)`` is a list of parameters. This list is populated as the pipeline runs, and CHiCAGO estimates them in turn.

A closer look at ``intData(cd)``:
```{r}
head(intData(cd), 2)
```

Columns:

* baitID: ID of baited fragment
* otherEndID: ID of other end fragment
* s_j: bait-specific scaling factor (Brownian noise)
* otherEndLen: The length of the other end fragment
* distSign: The distance from the baited fragment to the other end fragment. Positive and negative values indicate that the other end is, respectively, $5'$ and $3'$ of the baited fragment. NA indicates a trans interactions.
* isBait2Bait: TRUE if the other end fragment is also a baited fragment
* N.1, N.2, ...: Raw read counts per replicate (see ``?mergeSamples``).
* N: Merged count (see ``?mergeSamples``) or raw count in the case of single-replicate interaction calling.
* refBinMean: Can be ignored. (see ``?normaliseBaits``)
* s_i: other end-specific scaling factor (Brownian noise)
* NNb: "N normalised for baits", a count scaled up by accounting for s_j. May be useful for visualization.
* NNboe: "N normalised for baits and other ends"; may be useful for visualization.
* tlb: Class of other end, based on the number of fragments on other chromosomes that have read pairs.
* tblb: As tlb, for the bait fragment.
* Tmean: Expected count from technical noise.
* Bmean: Expected count from Brownian noise. (Thus, the expected count under the null hypothesis is Tmean + Bmean.)
* log.p: p-value associated with fragment pair, on log-scale.
* log.w: p-value weight, on log-scale.
* log.q: weighted p-value, on log-scale.
* score: Final CHiCAGO score.

**WARNING:** Many functions in CHiCAGO update ``intData(cd)`` by reference, which means that ``intData(cd)`` can change even when you do not explicitly assign to it.
To avoid this behaviour, copy the *chicagoData* object first:

```{r}
newCd = copyCD(cd)
```

##Session info

```{r}
sessionInfo()
```

##References
<!-- the bibliography automatically gets appended here -->
